

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the ???quantified self??? movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as ????????)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
 

##Loading and preprocessing ata data

```{r}  
activity <- read.csv("activity.csv")
summary(activity)
library(dplyr)
activity <- mutate(activity, date = as.Date(as.character(date), "%Y-%m-%d"))
```

##What is mean total number of steps taken per day?

 Missing values in the dataset can be ignored

1. Calculate the total number of steps taken per day
```{r}
# calculate total steps per day
totalsteps <- aggregate(steps ~ date, activity, sum )
# label the column
names(totalsteps)[names(totalsteps)=="steps"] <- "totalsteps"
# output
totalsteps
```
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
```{r}
# Calculate the total number of steps taken per day
takensteps <- aggregate(steps ~ date, activity, sum)
# histogram
hist(takensteps$steps, col="Green", ylab="Frequency", xlab="Total number of Steps",main="Total number of steps taken per day")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
# calcualte the mean of the each day
tendency <- aggregate(steps ~ date, activity, mean, na.rm=TRUE)
# # calcualte the median of the each day
medianTendency  <- aggregate(steps ~ date, activity, median, na.rm=TRUE)
# combine in one table
tendency <- cbind(tendency, medianTendency$steps)
# rename the mean column lable
names(tendency)[names(tendency)=="steps"] <- "mean"
# rename median column label 
names(tendency)[names(tendency)=="medianTendency$steps"] <- "median"
# print the table to the console
tendency 

# total daily mean and median averages respectively
mean(activity$steps, na.rm=TRUE)
median(activity$steps, na.rm=TRUE)
```

##What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
```{r}
# calculate mean steps by interval 
x <- aggregate(steps ~ interval, activity, mean, na.rm=TRUE)
# plot the line chart and label
plot(x, type="l", xlab="Interval", ylab="Average Steps Taken")
```  

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r}
# cross check the graph's largest interval
x$interval[x$steps==max(x$steps)]
```

##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  
```{r}
# Sum the NA values 
sum(is.na(activity$steps))
```  

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

The preponderance of missing data are complete days, impute no activity into the data set for missing values.  That is, impute a zero for NA values in the data set. 

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# create the new table to impute values for NA
imputed <- activity
# substitute NA with zero
imputed$steps[is.na(imputed$steps)] <- 0
# cross check, no NA's
sum(is.na(imputed$steps))
# cross check again, total entries add up to data set
sum(!is.na(imputed$steps))
```   

4. 
   a. Make a histogram of the total number of steps taken each day Calculate and report the mean and median total number of steps taken per day. Calculate and report the mean and median total number of steps taken per day.Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
# calculate the frequencies by day
imputedsum <- aggregate(steps ~ date, imputed, sum)
# Generate Histogram
hist(imputedsum$steps, col = "Green", ylab=" Frequency ", xlab="Activity Days")  

# calculate the mean by day 
tendency <- aggregate(steps ~ date, imputed, mean)
# calculate the median by day with the imputed data set
medianTendency  <- aggregate(steps ~ date, imputed, median)
# combine into on table
tendency <- cbind(tendency, medianTendency$steps)
# label the mean column 
names(tendency)[names(tendency)=="steps"] <- "mean"
# label the median column 
names(tendency)[names(tendency)=="medianTendency$steps"] <- "median"
# print out put to console
tendency 


# total daily mean and median averages respectively

mean(imputed$steps)
median(imputed$steps)
```

   

##Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.



1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  
```{r}
#create a vector of weekdays
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
#Use `%in%` and `weekdays` to create a logical vector
#Convert to `factor` and specify the `levels/labels`
imputed$partofweek <-  factor((weekdays(imputed$date) %in% weekdays1)+1L,
      levels=1:2, labels=c('weekend', 'weekday'))
#activity$partofweek
 aggregate(steps ~ partofweek, imputed, sum, na.rm=TRUE)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
# Make a two part plot
par(mfrow=c(2,1))
# get the weekday data
weekday <- subset(imputed, partofweek=="weekday")
# calculate the weekday mean
wd <- aggregate(steps ~ interval, weekday, mean, na.rm=TRUE)
# plot the average interval graph of weekdays
plot(wd, type="l", xlab="Weekday", ylab="Average Steps")

# get the weekday data 
weekend <- subset(imputed, partofweek=="weekend")
# calculate the weekday mean 
we <- aggregate(steps ~ interval, weekend, mean, na.rm=TRUE)
# plot the average interval graph of weekends
plot(we, type="l", xlab="Weekend", ylab="Average Steps")
